Appendix 4: Results

### 2016-03-14 Harris_2015/recovery.tex
final before resubmission to Ecology?

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Find all the csv files in the fakedata/matrices folder,}
\CommentTok{# then drop .csv}
\NormalTok{identifiers =}\StringTok{ }\KeywordTok{dir}\NormalTok{(}\StringTok{"fakedata/matrices"}\NormalTok{, }\DataTypeTok{pattern =} \StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.csv$"}\NormalTok{) %>%}
\StringTok{  }\KeywordTok{gsub}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.csv$"}\NormalTok{, }\StringTok{""}\NormalTok{, .)}

\KeywordTok{mclapply}\NormalTok{(identifiers, fit_all, }\DataTypeTok{mc.cores =} \NormalTok{mc.cores, }
         \DataTypeTok{mc.preschedule =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{mc.silent =} \OtherTok{TRUE}\NormalTok{) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{bind_rows}\NormalTok{() %>%}\StringTok{ }
\StringTok{  }\KeywordTok{write.csv}\NormalTok{(}\DataTypeTok{file =} \StringTok{"estimates.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\subsubsection{Logistic prior used for the Markov
network}\label{logistic-prior-used-for-the-markov-network}

@@

\includegraphics{recovery_files/figure-latex/unnamed-chunk-4-1.pdf}

@@

>>>>

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# Find all the csv files in the fakedata/matrices folder,}
\CommentTok{# then drop .csv}
\NormalTok{identifiers =}\StringTok{ }\KeywordTok{dir}\NormalTok{(}\StringTok{"fakedata/matrices"}\NormalTok{, }\DataTypeTok{pattern =} \StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.csv$"}\NormalTok{) %>%}
\StringTok{  }\KeywordTok{gsub}\NormalTok{(}\StringTok{"}\CharTok{\textbackslash{}\textbackslash{}}\StringTok{.csv$"}\NormalTok{, }\StringTok{""}\NormalTok{, .)}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{mclapply}\NormalTok{(identifiers, fit_all, }\DataTypeTok{mc.cores =} \NormalTok{mc.cores, }
         \DataTypeTok{mc.preschedule =} \OtherTok{FALSE}\NormalTok{, }\DataTypeTok{mc.silent =} \OtherTok{TRUE}\NormalTok{) %>%}\StringTok{ }
\StringTok{  }\KeywordTok{bind_rows}\NormalTok{() %>%}\StringTok{ }
\StringTok{  }\KeywordTok{write.csv}\NormalTok{(}\DataTypeTok{file =} \StringTok{"estimates.csv"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\subsection{Logistic prior used for the Markov
network}\label{logistic-prior-used-for-the-markov-network}

@@

@@

\includegraphics{recovery_files/figure-latex/unnamed-chunk-6-1.pdf}

\subsection{GLM coefficient
correlations}\label{glm-coefficient-correlations}

The GLM method produces two estimates for each species pair. These
estimates tend to be very similar, so most reasonable methods for
generating a consensus estimate should produce very similar results.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{get_glm_cor =}\StringTok{ }\NormalTok{function(identifier)\{}
  \NormalTok{data_filename =}\StringTok{ }\KeywordTok{paste0}\NormalTok{(}\StringTok{"fakedata/matrices/"}\NormalTok{, identifier, }\StringTok{".csv"}\NormalTok{)}

  \NormalTok{x =}\StringTok{ }\NormalTok{raw_obs[ , species_is_variable]}

  \NormalTok{coef_matrix =}\StringTok{ }\KeywordTok{matrix}\NormalTok{(}\DecValTok{0}\NormalTok{, }\KeywordTok{ncol}\NormalTok{(x), }\KeywordTok{ncol}\NormalTok{(x))}

  \CommentTok{# Fill in the coefficient matrix, column-by-column}
  \NormalTok{for(i in }\DecValTok{1}\NormalTok{:}\KeywordTok{ncol}\NormalTok{(x))\{}
    \NormalTok{if(}\KeywordTok{var}\NormalTok{(x[,i]) >}\StringTok{ }\DecValTok{0}\NormalTok{)\{}
      \NormalTok{coefs =}\StringTok{ }\KeywordTok{coef}\NormalTok{(}\KeywordTok{bayesglm}\NormalTok{(x[,i] ~}\StringTok{ }\NormalTok{x[ , -i], }\DataTypeTok{family =} \NormalTok{binomial))[-}\DecValTok{1}\NormalTok{]}
      \NormalTok{coef_matrix[i, -i] =}\StringTok{ }\NormalTok{coefs}
    \NormalTok{\}}
  \NormalTok{\}}

  \CommentTok{# Compare the estimates from the upper triangle with the estimates from }
  \CommentTok{# the estimates in the lower triangle (which becomes the upper triangle after }
  \CommentTok{# transposing)}
  \KeywordTok{cor}\NormalTok{(coef_matrix[}\KeywordTok{upper.tri}\NormalTok{(coef_matrix)], }\KeywordTok{t}\NormalTok{(coef_matrix)[}\KeywordTok{upper.tri}\NormalTok{(coef_matrix)])}
\NormalTok{\}}

\NormalTok{glm_correlations =}\StringTok{ }\KeywordTok{unlist}\NormalTok{(}\KeywordTok{mclapply}\NormalTok{(identifiers, get_glm_cor))}

\KeywordTok{mean}\NormalTok{(glm_correlations)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 0.9546387
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\KeywordTok{hist}\NormalTok{(glm_correlations)}
\end{Highlighting}
\end{Shaded}

\includegraphics{recovery_files/figure-latex/unnamed-chunk-7-1.pdf}

### 2016-03-14 Harris_2015/show-results.Rmd
final before resubmission to Ecology?

---
output: 
  pdf_document:
    keep_tex: true
title: "Appendix 4: Results"
subtitle: "Inferring species interactions from co-occurrence data with Markov networks"
author: David J. Harris
fontsize: 11pt
---

@@

```{r, cache = TRUE}
resids = function(data){
  resid(lm(truth ~ estimate + 0, data = data))
}

@@

@@

```{r, cache = TRUE}
landscape_estimates = x %>% 
  group_by(method, simulation_type) %>%
  do(data.frame(., resids = resids(.))) %>%
  ungroup %>%
  group_by(method, simulation_type, rep_name, n_sites) %>%
  summarise(r2 = 1 - sum(resids^2) / sum(truth^2))

@@

## Plot estimates versus "true" values across methods and lanscape sizes

The gray line indicates the 1:1 line, where models recover an unbiased estimate
of the "true" $\beta$ coefficients.  The GLM and Markov network methods estimate
the coefficients on the original scale, and approach unbiasedness as the number
of sites increases. The four methods based on (partial) correlations are
constrained to return values between -1 and 1, and do not return values on the
original scale.  The covariance and partial covariance (which are not scaled in
this way) may avoid this problem in future analyses (Loh and Wainwright 2013; 
see main text for full citation).

Note that the null model (Pairs) is not included in these graphs, as its
test statistic depends strongly on the size of the landscape and fills a range 
that would not fit on these graphs.

```{r, dev="png", dpi=500, cache = TRUE}
ggplot(filter(x, simulation_type == "no_env", method != "null"),
       aes(x = truth, y = estimate, color = factor(n_sites))) + 
  facet_wrap(~method) +
  geom_point(alpha = 0.01) +
  geom_abline(intercept = 0, slope = 1, size = 2, col = "darkgray") + 
  geom_smooth(method = "lm", fullrange = FALSE) + 
  coord_equal() + 
  scale_color_discrete(name = "n_sites")
```

@@

@@

pairs_summary = x[x$method == "null" & !grepl("pop", x$rep_name), ]
markov_summary = x[x$method == "Markov network" & !grepl("pop", x$rep_name), ]

@@

The `error_smoother` is a function that fits a generalized additive model to 
estimate the probability that a model estimate will match some criterion (e.g. 
statistical significance) for some parameter as a function of that parameter's
"true" value.  This function is used in Figure 4C and in calculating Type I
error rates below.

@@

error_smoother = function(data, f, values = truth_seq){
  predict(
    gam(
      f(data) ~ s(truth),
      data = data,
      family = binomial
    ),
    data.frame(truth = values),
    type = "response"
  )
}
```

@@

@@

# Calculate p(confidently wrong) for each model as a function of
# the true values
confidently_wrong = function(data){
  (data$sig_pos & data$truth < 0) | (data$sig_neg & data$truth > 0)
}
y_markov = error_smoother(markov_summary, confidently_wrong)
y_pairs = error_smoother(pairs_summary, confidently_wrong)

@@

plot(
  truth_seq,
  y_pairs,
  type = "l",
  xlab = "\"True\" interaction strength",
  ylab = "P(confidently predict wrong sign)",
  bty = "l",
  yaxs = "i",
  col = 2,
  ylim = c(0, .4),
  lwd = 2
)
mtext("C. Error rate vs.\ninteraction strength", side = 3, adj = 0, 
      font = 2, line = 1.2, cex = .9) 
lines(truth_seq, y_markov, lwd = 2)
legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), 
       col = c(2, 1), bty = "n")
dev.off()
```

@@

Type I error rates (probability of rejecting the null hypothesis given a 
"true" interaction strength of zero, according to the spline model defined by
`error_smoother` above). The smoother focuses on the 
`r markov_summary %>% summarize(mean(abs(truth) < 1)) %>% round(., 2) %>% multiply_by(100)`%
of the data points whose "true" values fall between -1 and 1, because its 
estimates near 0 are more accurate when they aren't affected by distant data 
points.

#### Markov network Type I error rates
```{r}
markov_summary %>% 
  filter(abs(truth) < 1) %>% 
  group_by(simulation_type) %>% 
  do(data.frame(
    `Type I error rate` = 
      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)
  )) %>%
  kable(digits = 3)
```

#### Null model Type I error rates
```{r}
pairs_summary %>% 
  filter(abs(truth) < 1) %>% 
  group_by(simulation_type) %>% 
  do(data.frame(
    `Type I error rate` = 
      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)
  )) %>%
  kable(digits = 3)
```

@@

The top and bottom 0.5% of the distribution have been omitted to prevent bad 
behavior by the smoother in the tails. The horizontal red line shows the nominal
95% coverage rate and the black vertical line marks where the "true" value
of $\beta$ was zero. The Type I error rates calculated above for the Markov 
network closely match (one minus) the values of these curves at zero.

@@

>>>>

---
output: 
  pdf_document:
    keep_tex: true
    toc: true
title: "Appendix 4: Results"
subtitle: "Inferring species interactions from co-occurrence data with Markov networks"
author: David J. Harris
fontsize: 11pt
---

@@

```{r, cache = FALSE}
resids = function(data){
  resid(lm(truth ~ estimate + 0, data = data))
}

@@

#### Plot the regressions used above to calculate R^2

The plots below each correspond to one of the regressions performed above to
calculate R^2 for a combination of simulation method and estimation method.
The null model ("Pairs") was not included in these plots because its Z-scores
were frequently too large to fit on the same scale as the other methods' estimates.
Below, the thinner dashed lines represent the 1:1 line, where the expected estimate
matches the "true" value, and the thicker solid lines represent a linear regression
through the origin, as calculated above. In general, the four correlation- or
partial correlation-based methods had slopes that were steeper than the 1:1 line,
because their estimates were constrained between -1 and 1. The GLM and Markov network
usually had slopes that were closer to 1.

(Warnings of the form "Removed N rows containing missing values
(geom_smooth)" can be ignored: they are simply noting that the regression lines 
were cut off in some panels of the figure).

```{r, dev="png", dpi=500, cache = TRUE}
plot_regressions = function(type, title){
  x[sample.int(nrow(x)) , ] %>%
  subset(simulation_type == type & method != "null") %>% 
  ggplot(aes(x = estimate, y = truth)) +
    facet_wrap(~method) +
    ylim(min(x$truth), max(x$truth)) + 
    geom_point(aes(color = factor(n_sites)), size = 0.2) + 
    geom_smooth(method = "lm", color = "black", formula = y ~ 0 + x, 
                fullrange = TRUE) + 
    geom_abline(intercept = 0, slope = 1, linetype = 2) +
    theme_bw() + 
    ggtitle(title)
}

plot_regressions("no_env", "'Baseline' regressions")
plot_regressions("env", "'Environment' regressions")
plot_regressions("abund", "'Abundance' regressions")
```

@@

```{r, cache = FALSE}
landscape_estimates = x %>% 
  group_by(method, simulation_type) %>%
  do(data.frame(., resids = resids(.))) %>%
  ungroup %>%
  group_by(method, simulation_type, rep_name, n_sites) %>%
  summarise(r2 = 1 - sum(resids^2) / sum(truth^2))

@@

@@

Note that, as mentioned in the main text, inferential statistics are only 
calculated for the `env` and `no_env` simulations (and not for the `abund`
simulations).

@@

```{r, cache = FALSE}

pairs_summary = x[x$method == "null" & x$simulation_type != "abund", ]
markov_summary = x[x$method == "Markov network" & x$simulation_type != "abund", ]

@@

The `error_smoother` is a function that fits a kernel smoother model to 
estimate the probability that a model estimate will match some criterion (e.g. 
statistical significance) for some parameter as a function of that parameter's
"true" value.  This function is used in Figure 4C and in calculating Type I
error rates below.

@@

error_smoother = function(data, f, values = truth_seq){
  out = ksmooth(
    x = data$truth,
    y = as.numeric(f(data)),
    x.points = values,
    kernel = "normal",
    bandwidth = 0.25
  )

  out$y[is.na(out$y)] = 0

  out
}
```

@@

```{r, cache = FALSE}

@@

# Calculate p(confidently wrong) for each model as a function of
# the true values
confidently_wrong = function(data){
  (data$sig_pos & data$truth < 0) | (data$sig_neg & data$truth > 0)
}
smoothed_markov = error_smoother(markov_summary, confidently_wrong)
smoothed_pairs = error_smoother(pairs_summary, confidently_wrong)

@@

plot(
  smoothed_pairs$x,
  smoothed_pairs$y,
  type = "l",
  xlab = "\"True\" interaction strength",
  ylab = "P(confidently predict wrong sign)",
  bty = "l",
  yaxs = "i",
  col = 2,
  ylim = c(0, .3),
  lwd = 2
)
mtext("C. Error rate vs.\ninteraction strength", side = 3, adj = 0, 
      font = 2, line = 1.2, cex = .9) 
lines(smoothed_markov$x, smoothed_markov$y, lwd = 2)
legend("topleft", lwd = 2, legend = c("Null model", "Markov network"), 
       col = c(2, 1), bty = "n")
dev.off()
```

@@

Type I error rates (probability of rejecting the null hypothesis given a 
"true" interaction strength of zero, according to the kernel smoother defined by
`error_smoother` above).

#### Markov network Type I error rates
```{r}
markov_summary %>% 
  filter(abs(truth) < 1) %>% 
  group_by(simulation_type) %>% 
  do(data.frame(
    `Type I error rate` = 
      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)$y
  )) %>%
  kable(digits = 3)
```

#### Null model Type I error rates
```{r}
pairs_summary %>% 
  filter(abs(truth) < 1) %>% 
  group_by(simulation_type) %>% 
  do(data.frame(
    `Type I error rate` = 
      error_smoother(., function(x){x$sig_pos | x$sig_neg}, values = 0)$y
  )) %>%
  kable(digits = 3)
```

@@

The top and bottom 0.5% of the distribution have been omitted to prevent bad 
behavior by the smoother (a generalized additive model) in the tails. The 
horizontal red line shows the nominal 95% coverage rate and the black vertical 
line marks where the "true" value of $\beta$ was zero. The Type I error rates 
calculated above for the Markov network closely match (one minus) the values of 
these curves at zero.

@@

## Plot power versus true $\beta$ value

Calculate p(confidently detect sign of $\beta$) as a function of
the "true"" values.  Pairs has a somewhat higher rate of rejecting the null 
hypothesis in the correct direction than the Markov network, but the benefit of 
this increase in power is more than offset by the large increases in Type I 
errors and sign errors demonstrated above and in Figure 4 of the main text.

```{r}
confidently_right = function(data){
  (data$sig_pos & data$truth > 0) | (data$sig_neg & data$truth < 0)
}

plot(error_smoother(markov_summary, confidently_right), 
     type = "l", 
     xlim = c(-5, 5),
     xlab = "'True' coefficient value",
     ylab = "Power",
     ylim = c(0, 1),
     yaxs = "i"
)
lines(error_smoother(pairs_summary, confidently_right), col = 2)
legend("topleft", legend = c("Markov network", "Pairs"), lty = 1, col = 1:2)
```
